Weakly Nonlinear MRI: Notes

### 2014-02-10 notes/mri_notes.tex
fixed sign error in viscous term

So instead, we begin with the momentum equation: 
%\begin{equation}
\[\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, = \, -\frac{1}{\rho}\nabla P \, - \, \nabla\Phi \, + \, \frac{1}{\rho} \left(\mathbf{J}\times\mathbf{B}\right) \, - \, \nu\nabla^2 \mathbf{u} \, - \, 2\mathbf{\Omega} \times \mathbf{u} \, - \, \mathbf{\Omega} \times \left(\mathbf{\Omega} \times \mathbf{r} \right)\]
%\end{equation}\label{momeqn}

@@

\begin{equation}
\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(v_r \mathbf{\hat{\phi}} - v_{\phi}\mathbf{\hat{r}}\right) \, = \, - r\left(\Omega^2 - \Omega_0^2\right)\mathbf{\hat{r}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  - \, \frac{1}{8 \pi \rho}\nabla B^2 + \frac{1}{4 \pi \rho} \left(\mathbf{B} \cdot \nabla\right)\mathbf{B}  \, - \, \nu\nabla^2 \mathbf{u} \,
\end{equation}

@@

And our momentum equation now reads: 
\[\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(u_x \mathbf{\hat{y}} - u_{y}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 x q \mathbf{\hat{x}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  - \, \frac{1}{8 \pi \rho}\nabla B^2 + \frac{1}{4 \pi \rho} \left(\mathbf{B} \cdot \nabla\right)\mathbf{B}  \, - \, \nu\nabla^2 \mathbf{u} \,\]

@@

\[\frac{\partial \mathbf{\widetilde{u}}}{\partial \widetilde{t}} \, + \, \mathbf{\widetilde{u}} \cdot \widetilde{\nabla} \mathbf{\widetilde{u}} \, + 2\Omega_0\left(\widetilde{u_x} \mathbf{\hat{y}} - \widetilde{u_{y}}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 \widetilde{x} q \mathbf{\hat{x}} \, -\frac{1}{\widetilde{\rho}}\widetilde{\nabla} \widetilde{P} \, +\frac{1}{\widetilde{\rho_b}}\widetilde{\nabla} \widetilde{P_b} \,  - \, \frac{1}{8 \pi \widetilde{\rho}}\widetilde{\nabla} \widetilde{B}^2 + \frac{1}{4 \pi \widetilde{\rho}} \left(\mathbf{\widetilde{B}} \cdot \widetilde{\nabla}\right)\mathbf{\widetilde{B}}  \, - \, \nu \widetilde{\nabla}^2 \mathbf{\widetilde{u}} \,\]

@@

%\begin{equation}
$\left(\Omega_0^2 r_0 \delta\right) \partial_t \mathbf{u} \, + \, \left(\Omega_0^2 r_0 \delta\right)\mathbf{u} \cdot \nabla \mathbf{u} \,+\, \left(\Omega_0^2 r_0 \delta\right) 2 u_x \mathbf{\hat{y}} \,-\, \left(\Omega_0^2 r_0 \delta\right) 2 u_y \mathbf{\hat{x}} \, = \, \left(\Omega_0^2 r_0 \delta\right) 2 x q \mathbf{\hat{x}} \,-\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho} \nabla P  \,+\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho_b} \nabla P_b \, - \, \frac{1}{8\pi}\Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right) \frac{1}{\beta} \frac{1}{\rho} \nabla B^2 \, + \, \frac{1}{4\pi} \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right) \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, - \, \nu\left(\frac{\Omega_0}{r_0 \delta}\right) \nabla^2 \mathbf{u}$ \\
%\end{equation}

@@

$\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + \, 2 u_x \mathbf{\hat{y}} - 2 u_y \mathbf{\hat{x}} \, = \, 2 x q \mathbf{\hat{x}} \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \nabla P \, + \,  \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \nabla P_b \, - \, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \nabla B^2 \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, - \, \frac{1}{R_e} \nabla^2 \mathbf{u}$ \\

@@

We will next take perturbations with respect to the background flow.

>>>>

So instead, we begin with the momentum equation: 
%\begin{equation}
\[\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, = \, -\frac{1}{\rho}\nabla P \, - \, \nabla\Phi \, + \, \frac{1}{\rho} \left(\mathbf{J}\times\mathbf{B}\right) \, + \, \nu\nabla^2 \mathbf{u} \, - \, 2\mathbf{\Omega} \times \mathbf{u} \, - \, \mathbf{\Omega} \times \left(\mathbf{\Omega} \times \mathbf{r} \right)\]
%\end{equation}\label{momeqn}

@@

\begin{equation}
\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(v_r \mathbf{\hat{\phi}} - v_{\phi}\mathbf{\hat{r}}\right) \, = \, - r\left(\Omega^2 - \Omega_0^2\right)\mathbf{\hat{r}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  - \, \frac{1}{8 \pi \rho}\nabla B^2 + \frac{1}{4 \pi \rho} \left(\mathbf{B} \cdot \nabla\right)\mathbf{B}  \, + \, \nu\nabla^2 \mathbf{u} \,
\end{equation}

@@

And our momentum equation now reads: 
\[\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(u_x \mathbf{\hat{y}} - u_{y}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 x q \mathbf{\hat{x}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  - \, \frac{1}{8 \pi \rho}\nabla B^2 + \frac{1}{4 \pi \rho} \left(\mathbf{B} \cdot \nabla\right)\mathbf{B}  \, + \, \nu\nabla^2 \mathbf{u} \,\]

@@

\[\frac{\partial \mathbf{\widetilde{u}}}{\partial \widetilde{t}} \, + \, \mathbf{\widetilde{u}} \cdot \widetilde{\nabla} \mathbf{\widetilde{u}} \, + 2\Omega_0\left(\widetilde{u_x} \mathbf{\hat{y}} - \widetilde{u_{y}}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 \widetilde{x} q \mathbf{\hat{x}} \, -\frac{1}{\widetilde{\rho}}\widetilde{\nabla} \widetilde{P} \, +\frac{1}{\widetilde{\rho_b}}\widetilde{\nabla} \widetilde{P_b} \,  - \, \frac{1}{8 \pi \widetilde{\rho}}\widetilde{\nabla} \widetilde{B}^2 + \frac{1}{4 \pi \widetilde{\rho}} \left(\mathbf{\widetilde{B}} \cdot \widetilde{\nabla}\right)\mathbf{\widetilde{B}}  \, + \, \nu \widetilde{\nabla}^2 \mathbf{\widetilde{u}} \,\]

@@

%\begin{equation}
$\left(\Omega_0^2 r_0 \delta\right) \partial_t \mathbf{u} \, + \, \left(\Omega_0^2 r_0 \delta\right)\mathbf{u} \cdot \nabla \mathbf{u} \,+\, \left(\Omega_0^2 r_0 \delta\right) 2 u_x \mathbf{\hat{y}} \,-\, \left(\Omega_0^2 r_0 \delta\right) 2 u_y \mathbf{\hat{x}} \, = \, \left(\Omega_0^2 r_0 \delta\right) 2 x q \mathbf{\hat{x}} \,-\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho} \nabla P  \,+\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho_b} \nabla P_b \, - \, \frac{1}{8\pi}\Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right) \frac{1}{\beta} \frac{1}{\rho} \nabla B^2 \, + \, \frac{1}{4\pi} \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right) \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, + \, \nu\left(\frac{\Omega_0}{r_0 \delta}\right) \nabla^2 \mathbf{u}$ \\
%\end{equation}

@@

$\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + \, 2 u_x \mathbf{\hat{y}} - 2 u_y \mathbf{\hat{x}} \, = \, 2 x q \mathbf{\hat{x}} \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \nabla P \, + \,  \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \nabla P_b \, - \, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \nabla B^2 \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, + \, \frac{1}{R_e} \nabla^2 \mathbf{u}$ \\

@@

We will next take perturbations with respect to the background flow. The steady solution is $\mathbf{u} = -2 q \Omega_0 \mathbf{\hat{y}}$, $P = P_b$, $\rho = \rho_b$, and $\mathbf{B} = B_0 \mathbf{\hat{z}}$, so we find perturbation solutions of the following form (1 subscripts denote perturbations): \\

$\mathbf{u} \, = \, -q \Omega_0 x \mathbf{\hat{y}} \, + \, \mathbf{u_1}$ \\

$\rho \, = \, \rho_b + \rho_1$ \\

$P \, = \, P_b \, + \, P_1$ \\

$\mathbf{B} \, = \, B_0\mathbf{\hat{z}} + \mathbf{B_1}$ \\

We will begin with the momentum equation. To make this more explicit, we write out all 3 equations (pre-perturbation): \\

$\partial_t u_x \, + \, \left(\mathbf{u} \cdot \nabla\right)u_x \, - \, 2 u_y \, = \, 2 x q \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_x P \, + \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \partial_x P_b \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_x \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_x \, + \, \frac{1}{R_e}\nabla^2 u_x  $ \\

$\partial_t u_y \, + \, \left(\mathbf{u} \cdot \nabla\right)u_y \, + \, 2u_x =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_y P \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_y \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_y \, + \, \frac{1}{R_e}\nabla^2 u_y  $ \\

$\partial_t u_z \, + \, \left(\mathbf{u} \cdot \nabla\right)u_z \,  =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_z P \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_z \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_z \, + \, \frac{1}{R_e}\nabla^2 u_z $ \\

Where we have made the substitution $\partial_y P_b = \partial_y \rho_b = 0$, as the steady barotropic solution must be axisymmetric. \\

These become: \\

$\partial_t u_{1x} \, + \, \left(\mathbf{u} \cdot \nabla\right)u_{1x} \, + \, 2 q \Omega_0 x \, - \, 2 u_{1y}\,= \, 2 x q \, - \, \frac{\epsilon^2}{\delta^2} \left( \frac{\partial_x P_1}{\rho_1 + \rho_b} - \frac{\rho_1\partial_x P_b}{\rho_b^2 + \rho_1\rho_b}\right) \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_x \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_x \, + \, \frac{1}{R_e}\nabla^2 u_x  $ \\
